A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-10-10 manuscript.qmd
small comment added
extended methods

The mixture model assumes that the data is generated by a set of $M$ different underlying models. In our case, we introduce $M$ distinct lymphatic spread models. Each patient $k$ is assumed to be generated by one specific model from this set of $M$ models.

@@

The latent variables $\boldsymbol{\epsilon}_k$ are unknown and must be inferred from the observed data. The mixture model assumes that each patient is generated by exactly one of the $M$ models. However, in our application, this assumption does not reflect continuous or overlapping behaviors between subsites. Instead, we infer a probabilistic distribution over the latent variables, given the observed data and model parameters. The posterior distribution:

@@

@@

--> Add predictions and compare to the prevalence. Additionally we could compare to tumor location predictions.

>>>>

The mixture model assumes that the data is generated by a set of $M$ different lymphatic spread models. Each patient $k$ is then assumed to be generated by one specific model from this set of $M$ models.

@@

The latent variables $\boldsymbol{\epsilon}_k$ are unknown and must be inferred from the observed data. The mixture model assumes that each patient is generated by exactly one of the $M$ models. To do so a probabilistic distribution over the latent variables, given the observed data and model parameters is computed, the posterior distribution:

@@

It is important to mention, that the latent variables $\boldsymbol{\epsilon}$ are only used for parameter estimation. The subsite specific mixture parameters $\boldsymbol{\pi}^s$ are the only parameters which are used for later risk predictions. In @fig-model-simple the mixture parameters are illustrated. Subsites with different spread patterns, such as Gum (C03) and Base of tongue (C01), are expected to get different model assignments. Nonetheless, the latent variables for two patients with the same diagnosis, i.e. same involvement $\boldsymbol{\xi}$ and and t-stage $T$, but different subsites are the same.

![Illustration of mixture parameter assignment. Since Gum and Base of tongue express different spread patterns, the two models are expected to have different model assignments. The arrow visibility represents the value of the mixture parameter $\pi$, where the more visible the arrow, the larger the value for $\pi$](figures/mixture_model_simplified.png){#fig-model-simple}

@@

add some analysis (level specific predictions and also comparison to single tumor location models.)
